---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(opendpd)
```

# opendpd

<!-- badges: start -->
<!-- badges: end -->

The goal of opendpd is to provide an easy-to-use R interface for accessing various Dallas Police public safety datasets via the Socrata Open Data API, including functions for data retrieval and basic cleaning.

## Installation

You can install the development version of opendpd like so:

``` r
# install.packages("remotes") # Uncomment if user might not have remotes
remotes::install_github("Steal-This-Code/opendpd")
```

## Usage
```{recommended libraries}
# Optional: Load other packages needed for examples here if necessary
# library(dplyr)
# library(lubridate)

```

## Fetching Data

The package provides functions starting with `get_` to retrieve data from different datasets. They handle API pagination automatically and allow filtering via dedicated arguments or a custom `where` clause.

## Police Incidents

This is the primary dataset for reported crime incidents.

```{r example}
# Get the first 50 incidents in the Central division for March 2024
incidents_mar_central <- get_incidents(
  start_date = "2024-03-01",
  end_date = "2024-03-31",
  division = "CENTRAL",
  limit = 50
)
print(head(incidents_mar_central))

# Get incidents as an sf object (requires sf package)
# Note: Uses State Plane coordinates (EPSG:2276)
# incidents_sf <- get_incidents(limit = 10, convert_geo = TRUE)
# print(class(incidents_sf))

```

## Police Arrests

Retrieve data on arrests made by Dallas Police.

```{r example 2}
# Get arrests for specific beats in March 2024
arrests_mar_beats <- get_arrests(
  start_date = "2024-03-01",
  end_date = "2024-03-31",
  beat = c(114, 121), # Note: Uses 'arlbeat' field from API
  limit = 25
)
print(head(arrests_mar_beats))

```

## Arrest Charges

Retrieve details about charges associated with arrests.

```{r example 3}
# Get Felony charges during April 2024
charges_apr_felony <- get_charges(
  start_date = "2024-04-01",
  end_date = "2024-04-30",
  severity = "FELONY",
  limit = 50
)
print(head(charges_apr_felony))

```

## Use of Force

Retrieve Response to Resistance / Use of Force data. Requires specifying the year, as data is stored in separate yearly datasets (currently supports 2017-2020). Column names may vary by year.

```{r example 4}
# Get UoF incidents for 2020 where force type involved 'Physical Force'
uof_2020_physical <- get_uof(
  year = 2020,
  force_type = "Physical Force",
  limit = 25
)
print(head(uof_2020_physical))

```

## Exploring Filter Values

Before filtering with the `get_` functions, you might want to see the available options for certain fields.

```{r example 5}
# See distinct divisions in the Incidents dataset
incident_divisions <- list_distinct_values(field = "division", dataset = "incidents")
print(incident_divisions)

# See distinct severities in the Charges dataset
charge_severities <- list_distinct_values(field = "severity", dataset = "charges")
print(charge_severities)

# See distinct penalty classes ('pclass') from Charges data
charge_pclasses <- list_distinct_values(field = "pclass", dataset = "charges")
print(charge_pclasses)

# See distinct arrest districts ('arldistrict') from Arrests data
arrest_districts <- list_distinct_values(field = "arldistrict", dataset = "arrests")
print(arrest_districts)

```

## Cleaning Data

The package includes functions to perform basic cleaning (text formatting, date parsing) and standardization for specific datasets. These are typically used after retrieving the data.

## Basic Cleaning (Text and Dates)

Functions like `clean_incidents_data()` and `clean_arrests_data()` apply standard text cleaning (lowercase, trim whitespace) to common fields and parse date/time columns into POSIXct objects (using `lubridate`, requires installation).

```{r example 6}
# Fetch raw incident data
raw_incidents <- get_incidents(limit = 20, start_date = "2024-04-01")

# Apply cleaning
# Needs stringr & lubridate: install.packages(c("stringr", "lubridate"))
cleaned_incidents <- clean_incidents_data(raw_incidents)

# Check date class and cleaned division
print(class(cleaned_incidents$date1))
print(head(cleaned_incidents$division))

# You can also clean arrest data
# raw_arrests <- get_arrests(limit=20)
# cleaned_arrests <- clean_arrests_data(raw_arrests)
```

## Standardizing Categorical Fields

Functions like `standardize_division()` and `standardize_district()` map variations in specific fields (like police division or council district in the incidents data) to a standard set of categories and convert them to factors. This is useful for analysis after basic cleaning.

```{r example 7}
# Assuming 'cleaned_incidents' from the previous step
standardized_incidents <- standardize_division(cleaned_incidents)
standardized_incidents <- standardize_district(standardized_incidents) # Can chain them

# See the standardized values and factor levels
print(table(standardized_incidents$division, useNA = "ifany"))
print(levels(standardized_incidents$district))

```

## Data Sources

Data is sourced from the [Dallas Open Data Portal](https://www.dallasopendata.com/). Specific datasets used include Police Incidents, Arrests, Arrest Charges, and Response to Resistance (Use of Force). Field names and data availability are subject to change based on the source portal. 

All data are made available through the [Open Data Commons Attribution License](http://opendatacommons.org/licenses/by/1.0/)
