---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# opendpd

<!-- badges: start -->
<!-- badges: end -->

The goal of opendpd is to provide an easy-to-use R interface for accessing various Dallas Police public safety datasets via the Socrata Open Data API, including functions for data retrieval and basic cleaning.

## Installation

You can install the development version of opendpd like so:

``` r
# install.packages("remotes") # Uncomment if user might not have remotes
remotes::install_github("Steal-This-Code/opendpd")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(opendpd)

# Get the 10 most recent police incidents
recent_incidents_raw <- get_incidents(limit = 10, `$order` = "date1 DESC")

# Apply basic cleaning (lowercase text, parse dates)
# Needs stringr and lubridate installed
recent_incidents_clean <- clean_incidents_data(recent_incidents_raw)

# Show the first few rows and structure
print(head(recent_incidents_clean))
str(recent_incidents_clean)
